Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain

### 2018-10-16 pkg/disttree/inst/draft/draft.Rnw
some more revision based on reviews and feedback
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2327 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: c4f83dac6f05d1a0d58f5b78b00dd29d8faf8b49

In weather forecasting, these properties are especially appealing in 
mountainous regions and complex terrain where a wide range of local-scale
effects are not yet resolved by the NWP models. 
Thus, effects with abrupt changes and possibly nonlinear interactions 
might be required to account for site-specific unresolved features.
To illustrate this in practice,
precipitation forecasts are obtained with distributional forests at 95
meteorological stations in a mountainous region in the Alps, covering mainly
Tyrol, Austria, and adjacent areas (see the map in Figure~\ref{fig:all-map}).
More specifically, a Gaussian distribution left-censored at zero, is employed to model 24-hour total precipitation so that the
zero-censored point mass describes the probability of observing no precipitation 
on a given day (see Figure~\ref{fig:axams-dist}). Forecasts for July are
established based on data from the same month over the years 1985--2012 including
80~covariates derived from a wide range of different NWP quantities. As
Figure~\ref{fig:axams-dist} shows, the station-wise forests yield a full
distributional forecast for each day -- here for one specific day (July~24) at one
station (Axams) over four years (2009--2012) -- based on the previous 24~years
as learning data. The corresponding observations conform reasonably well with
the predictions. In Section~\ref{sec:precipitation} we investigate the
performance of distributional forests in this forecasting task in more detail.
Compared to three alternative zero-censored Gaussian models distributional forests
perform at least on par and sometimes clearly better while requiring no meteorological
knowledge about the athmospheric processes which drive formation of precipitation
for the model specification. The three alternatives are: a standard ensemble model
output statistics approach \citep[EMOS,][]{Gneiting+Raftery+Westveld:2005} based
on an NGR, a GAMLSS with regressors prespecified based on meteorological expertise
\citep[following][]{Stauffer+Umlauf+Messner:2017}, and a boosted GAMLSS
\citep{Hofner+Mayr+Schmid:2016} using non-homogeneous boosting
\citep{Messner+Mayr+Zeileis:2017} as an alternative technique for variable selection among all
80~available regressors.

@@

Many statistical weather forecasting models leverage the strengths of modern
numerical weather prediction (NWP) systems \citep[see][]{Bauer+Thorpe+Brunet:2015}.  
One frequently used method based on distributional regression models is the 
ensemble model output statistics (EMOS) approach first proposed by
\cite{Gneiting+Raftery+Westveld:2005} to produce high-quality forecasts for
specific quantities and sites.  
\hlc{In case of precipitation forecasting EMOS typically uses the ensemble mean of 
``total precipitation'' (\emph{tp}) forecasts as predictor for the location 
parameter $\mu$ and the corresponding ensemble standard deviation for the 
scale parameter $\sigma$ of the statistical model based on the assumption
of the response following a Gaussian distribution left-censored at zero.
This distributional approach of modeling both parameters allows to correct 
for possible errors of the NWP ensemble in both, the expectation but also the 
uncertainty of a specific forecast. In order to assure positivity of the 
scale parameter a log-link is applied. Hence, the two distribution parameters 
are modeled by the linear functions} 
$\mu = \beta_0 + \beta_1 \cdot \mbox{mean}(\mbox{tp}^{\frac{1}{p}})$ and  
$\log(\sigma) = \gamma_0 + \gamma_1 \cdot \mbox{sd}(\mbox{tp}^{\frac{1}{p}})$
with power transformation parameter $\frac{1}{p}$ and 
regression coefficients $\beta_0$, $\beta_1$, $\gamma_0$ and $\gamma_1$,  
\hlc{hereafter referred to as a basic EMOS model.}

\hlc{\textit{FIXME: notation with or withoug power transformation parameter?}}

While this approach alone is already highly effective in the plains,
it typically does not perform as well in complex terrain due to unresolved
effects in the NWP system. For example, in the Tyrolean Alps -- considered
in the following case study -- the NWP grid cells of $50 \times 50$
km$^2$ are too coarse to capture single mountains, narrow valleys, etc. 
Therefore, it is often possible to substantially improve the predictive 
performance of the EMOS by including additional predictor variables, either 
from local meteorological observations or an NWP model. Unfortunately, it is typically unknown 
which variables are relevant for improving the predictions. Simply including
all available variables may be computationally burdensome and
can lead to overfitting but, on the other hand, excluding too many variables
may result in a loss of valuable information. Therefore, selecting
the relevant variables and interactions among all possible covariates is 
crucial for improving the statistical forecasting model.

@@

\subsubsection*{Variable Selection}
\hlc{Generally, there are many possibilities how to specify the variables that are
to be included in a model. Especially for a low number of variables all types
of interactions can be employed and mixed effect models and separations into 
main effects and interaction effects can be considered allowing for easier 
interpretations} (see for example \cite{Wood+Scheipl+Faraway:2013}, 
\cite{Goicoa+Adin+Ugarte:2018}, \cite{Ugarte+Adin+Goicoa:2017}).
\hlc{However, in the application presented in} Section~\ref{sec:precipitation}
\hlc{alltogether 80 variables are available. Even if only paired interactions were
considered this would already lead to 3160 additional covariates in the model
making it very difficult to keep the model manageable regarding computational
costs but also interpretability. Therefore, only main effects are considered
for the distributional forest and the boosted GAMLSS which both automatically
select covariates while meteorological expert knowledge was applied to 
select covariates and specify interactions for the prespecified GAMLSS.}

@@

\subsubsection{Distributional specifications}
%% choice of distribution mit Verweis auf Supplement distributions
\hlc{Choosing an adequate distributional family is a crucial step and forms
the basis required to build a well fitting model. Therefore, this
decision should be made thoughtfuly. 
In the presented application we applied a Gaussian distribution left-censored 
at zero as this has proven to be a well working choice in
earlier literature} \citep{Stauffer+Mayr+Messner:2017}. 
\hlc{To test for robustness against distributional misspecifation 
different distributional assumptions have been considered as well. 
In Supplement 1 (Different Response Distributions) the exact
same evaluations as in} Section~\ref{sec:all} \hlc{are carried out for 15
observation stations, once employing a Gaussian hurdle model and once
with a logistic distribution left-censored at zero in order to account
for heavier tails. Under both assumptions the obtained results 
are similar to those presented in the paper applying a Gaussian distribution
left-censored at zero. However, comparing the two censored distributions,
at many stations the models perform not as well under the logisctic distribution 
as under the Gaussian distribution. The hurdle model leads to slight improvements 
in some settings, but this comes with more variablity (especially for EMOS and the 
boosted GAMLSS).
Alltogether, the distributional forest has proven to be very robust against
distributional misspecifications.}

@@

The fitted distributional forest for July 24 and
observation station Axams (including Figure~\ref{fig:axams-dist})
is reproducible using 
\code{demo("RainAxams", package = "disttree")}.
This also includes fitting the other zero-censored
Gaussian models considered in this paper and generating the corresponding 
QQ~plots (Figure~\ref{fig:axams-qq}) and
PIT histograms (Figure~\ref{fig:axams-pit}).
Full replication of all results can be obtained with
\code{demo("RainTyrol", package = "disttree")}
requiring the companion \textsf{R} package \textbf{RainTyrol} (version~\Sexpr{packageVersion("RainTyrol")}),
also available within the \textsf{R}-Forge~project.

>>>>

In weather forecasting, these properties are especially appealing in 
mountainous regions and complex terrain where a wide range of local-scale
effects are not yet resolved by the NWP models. 
Thus, effects with abrupt changes and possibly nonlinear interactions 
might be required to account for site-specific unresolved features.
To illustrate this in practice,
precipitation forecasts are obtained with distributional forests at 95
meteorological stations in a mountainous region in the Alps, covering mainly
Tyrol, Austria, and adjacent areas (see the map in Figure~\ref{fig:all-map}).
More specifically, a Gaussian distribution left-censored at zero, is employed to model 24-hour total precipitation so that the
zero-censored point mass describes the probability of observing no precipitation 
on a given day (see Figure~\ref{fig:axams-dist}). Forecasts for July are
established based on data from the same month over the years 1985--2012 including
80~covariates derived from a wide range of different NWP quantities. As
Figure~\ref{fig:axams-dist} shows, the station-wise forests yield a full
distributional forecast for each day -- here for one specific day (July~24) at one
station (Axams) over four years (2009--2012) -- based on the previous 24~years
as learning data. The corresponding observations conform reasonably well with
the predictions. In Section~\ref{sec:precipitation} we investigate the
performance of distributional forests in this forecasting task in more detail.
Compared to three alternative zero-censored Gaussian models distributional forests
perform at least on par and sometimes clearly better while requiring no meteorological
knowledge about the atmospheric processes which drive formation of precipitation
for the model specification. The three alternatives are: a standard ensemble model
output statistics approach \citep[EMOS,][]{Gneiting+Raftery+Westveld:2005} based
on an NGR, a GAMLSS with regressors prespecified based on meteorological expertise
\citep[following][]{Stauffer+Umlauf+Messner:2017}, and a boosted GAMLSS
\citep{Hofner+Mayr+Schmid:2016} using non-homogeneous boosting
\citep{Messner+Mayr+Zeileis:2017} as an alternative technique for variable selection among all
80~available regressors.

@@

Many statistical weather forecasting models leverage the strengths of modern
numerical ensemble prediction systems (EPSs; see \cite{Bauer+Thorpe+Brunet:2015}).
\hlc{EPSs not only predict the most likely future state of the atmosphere but provide
information about the uncertainty for a specific quantity and weather situation.
This is done by running the NWP model several times using slightly perturbed 
initial conditions and model specifications to account for uncertainties in both,
the initial state and the model and its parametrizations.
One frequently used method based on distributional regression models is the 
ensemble model output statistics (EMOS) approach first proposed by}
\cite{Gneiting+Raftery+Westveld:2005} \hlc{to produce high-quality forecasts for
specific quantities and sites.  
In case of precipitation forecasting EMOS typically uses the ensemble mean of 
``total precipitation''} (\emph{tp}) \hlc{forecasts as predictor for the location 
parameter $\mu$ and the corresponding ensemble standard deviation for the 
scale parameter $\sigma$ of the statistical model based on the assumption
of the response following a Gaussian distribution left-censored at zero.
This distributional approach of modeling both parameters allows to correct 
for possible errors of the NWP ensemble in both, the expectation but also the 
uncertainty of a specific forecast. In order to assure positivity of the 
scale parameter a log-link is applied. Hence, the two distribution parameters 
are modeled by the linear functions} 
$\mu = \beta_0 + \beta_1 \cdot \emph{mean}(\emph{tp})$ and  
$\log(\sigma) = \gamma_0 + \gamma_1 \cdot \emph{sd}(\emph{tp})$
with regression coefficients $\beta_0$, $\beta_1$, $\gamma_0$ and $\gamma_1$,  
\hlc{hereafter referred to as a basic EMOS model} \citep{Gebetsberger+Messner+Mayr:2017}.
%$\mu = \beta_0 + \beta_1 \cdot \mbox{mean}(\mbox{tp}^{\frac{1}{p}})$ and  
%$\log(\sigma) = \gamma_0 + \gamma_1 \cdot \mbox{sd}(\mbox{tp}^{\frac{1}{p}})$
%with power transformation parameter $\frac{1}{p}$ and 
%regression coefficients $\beta_0$, $\beta_1$, $\gamma_0$ and $\gamma_1$,  
%\hlc{hereafter referred to as a basic EMOS model \citep{Gebetsberger+Messner+Mayr:2017}.}

\hlc{\textit{FIXME: notation with or without power transformation parameter?}}

While this approach alone is already highly effective in the plains,
it typically does not perform as well in complex terrain due to unresolved
effects in the NWP system \citep{Bauer+Thorpe+Brunet:2015}. 
For example, in the Tyrolean Alps -- considered
in the following case study -- the NWP grid cells of $50 \times 50$
km$^2$ are too coarse to capture single mountains, narrow valleys, etc. 
Therefore, it is often possible to substantially improve the predictive 
performance of the EMOS by including additional predictor variables, either 
from local meteorological observations or an NWP model. Unfortunately, it is typically unknown 
which variables are relevant for improving the predictions. Simply including
all available variables may be computationally burdensome and
can lead to overfitting but, on the other hand, excluding too many variables
may result in a loss of valuable information. Therefore, selecting
the relevant variables and interactions among all possible covariates is 
crucial for improving the statistical forecasting model.

@@

\subsubsection*{Variable Selection}
\hlc{Generally, there are many possibilities how to specify the variables that are
to be included in a model. Especially for a low number of variables all types
of interactions can be employed and mixed effect models and separations into 
main effects and interaction effects can be considered allowing for easier 
interpretations} (see for example \cite{Wood+Scheipl+Faraway:2013}, 
\cite{Goicoa+Adin+Ugarte:2018}, \cite{Ugarte+Adin+Goicoa:2017}).
\hlc{However, in the application presented in} Section~\ref{sec:precipitation}
\hlc{altogether 80 variables are available. Even if only paired interactions were
considered this would already lead to 3160 additional covariates in the model
making it very difficult to keep the model manageable regarding computational
costs but also interpretability. Therefore, only main effects are considered
for the distributional forest and the boosted GAMLSS which both automatically
select covariates while meteorological expert knowledge was applied to 
select covariates and specify interactions for the prespecified GAMLSS.}

@@

\subsubsection{Distributional specifications}
%% choice of distribution mit Verweis auf Supplement distributions
\hlc{Choosing an adequate distributional family is a crucial step and forms
the basis required to build a well fitting model. Therefore, this
decision should be made thoughtfully. 
In the presented application we applied a Gaussian distribution left-censored 
at zero as this has proven to be a well working choice in
earlier literature} \citep{Stauffer+Mayr+Messner:2017}. 
\hlc{To test for robustness against distributional misspecifation 
different distributional assumptions have been considered as well. 
In Supplement 1 (Different Response Distributions) the exact
same evaluations as in} Section~\ref{sec:all} \hlc{are carried out for 15
observation stations, once employing a Gaussian hurdle model and once
with a logistic distribution left-censored at zero in order to account
for heavier tails. Under both assumptions the obtained results 
are similar to those presented in the paper applying a Gaussian distribution
left-censored at zero. However, comparing the two censored distributions,
at many stations the models perform not as well under the logistic distribution 
as under the Gaussian distribution. The hurdle model leads to slight improvements 
in some settings, but this comes with more variability (especially for EMOS and the 
boosted GAMLSS).
Altogether, the distributional forest has proven to be very robust against
distributional misspecifications.}

@@

The fitted distributional forest for July 24 and
observation station Axams (including Figure~\ref{fig:axams-dist})
is reproducible using 
\code{demo("RainAxams", package = "disttree")}.
This also includes fitting the other zero-censored
Gaussian models considered in this paper and generating the corresponding 
QQ~plots (Figure~\ref{fig:axams-qq}).
Full replication of all results can be obtained with
\code{demo("RainTyrol", package = "disttree")}
requiring the companion \textsf{R} package \textbf{RainTyrol} (version~\Sexpr{packageVersion("RainTyrol")}),
also available within the \textsf{R}-Forge~project.

### 2018-10-16 pkg/disttree/inst/draft/rejoinder.tex
some more revision based on reviews and feedback
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2327 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: c4f83dac6f05d1a0d58f5b78b00dd29d8faf8b49

\begin{reto}
Wo baut man das am besten ein?
\end{reto}

@@

\textit{TO DO}

\begin{reto}
Es gibt da ein klassisches MOS-Paper in welchem Postprocessing
mit einem Globalmodell und einem Lokalmodell gemacht werden.
Das ist nicht die Referenz die der Reviewer will.
Das Problem ist, dass es explizit dazu nix gibt. Das ist einfach
jedem von uns klar. M\"uller zeigt aber in seinen Figures (f\"ur
Wind und Temperatur), dass ein Modell mit h\"oherer Aufl\"osung
einen geringeren Fehler aufweist und sagt folgendes:

``\textit{The largest differences are found in the raw
model forecast for stations in complex terrain. For wind
speed, the increase in resolution is more beneficial than
for temperature but is again negligible for MOS forecasts.
[\dots]
A higher resolution will
further reduce the model bias, especially in complex
terrain, but this can also be achieved with statistical
postprocessing if observations are available. Furthermore,
a postprocessed temperature forecast at 12-km
resolution outperforms a raw forecast at 3-km, regardless
of the complexity of the terrain}''.

\begin{verbatim}
@article{mueller2011,
    author = {M{\"u}ller, M. D.},
    title = {Effects of Model Resolution and Statistical Postprocessing on Shelter Temperature and Wind Forecasts},
    journal = {Journal of Applied Meteorology and Climatology},
    volume = {50},
    number = {8},
    pages = {1627--1636},
    year = {2011},
    doi = {10.1175/2011JAMC2615.1},
}
\end{verbatim}

Oder Meyers and Steenburgh (2013) der aufzeigt, dass auch der Mensch sich mit
der Wettervorhersage im Gebirge schwer tut bzw. dort die groessten Challenges
zu finden sind mit all den lokalen und kleinraeumigen Eigenschaften und Einfluessen.
Habe das Buch grad nicht rumliegen, ist wohl aber auch nicht zwingend notwendig.

\begin{verbatim}
@Inbook{meyers2013,
    author = {Meyers, Michael P. and Steenburgh, W. James},
    editor = {Chow, Fotini K. and De Wekker, Stephan F.J. and Snyder, Bradley J.},
    title = {Mountain Weather Prediction: Phenomenological Challenges and Forecast Methodology},
    bookTitle = {Mountain Weather Research and Forecasting: Recent Progress and Current Challenges},
    year = {2013},
    publisher = {Springer Netherlands},
    pages = {1--34},
    isbn = {978-94-007-4098-3},
    doi = {10.1007/978-94-007-4098-3_1},
}
\end{verbatim}
\end{reto}

@@

\begin{center}
\includegraphics[width = 0.7\textwidth]{varimp_gb.jpeg}
\end{center}

@@

>>>>

Ensemble models are now introduced in more detail in the
first paragraph of Section~3.

@@

This is itself a complex topic with an own scientific community
focusing on the development and improvement of numerical weather
forecasts. Bauer, Thorphe, and Brunet (2015) give a good overview
of the history and open scientific challenges of NWP ensemble forecast
models including a discussion about unresolved processes and possible
solutions beyond the current state of NWP models.
We have added this reference next to the statement you are referring to.

The following figure shows a comparison of the true local topography
(Robinson et~al. 2014; left)
and the model topography of the GEFS reforecast model
($0.5^\circ \times 0.5^\circ$, approx. $50 \times 50~km^2$; right) to illustrate
the extent of simplification needed in such global weather forecast models.

\begin{center}
\includegraphics[width = .8\textwidth]{topo-GFS.pdf}
\end{center}

%\begin{verbatim}
%@Article{Bauer2015,
%    author   = {Bauer, Peter and Thorpe, Alan and Brunet, Gilbert},
%    title    = {The Quiet Revolution of Numerical Weather Prediction},
%    journal  = {Nature},
%    year     = {2015},
%    month    = {Sep},
%    day      = {02},
%    volume   = {525},
%    pages    = {47--55},
%    doi      = {10.1038/nature14956}
%}
%\end{verbatim}
%Bauer and Thorpe (2015, p.~52): ``\textit{The scientific challenges:
%[\dots] the uncertainties inherent to physical parameterizations, either
%from incomplete process understanding or the dilemma of representing
%the impact of unresolved processes on the resolved scales, may require a
%fundamentally different approach.''

@@

\begin{center}
\includegraphics[width = .7\textwidth]{varimp_gb.pdf}
\end{center}

@@

\item[] M{\"u}ller, M. D. (2011): Effects of Model Resolution and Statistical
Postprocessing on Shelter Temperature and Wind Forecasts,
\textit{Journal of Applied Meteorology and Climatology} 
\textbf{50 8}: 1627--1636.
%doi:10.1175/2011JAMC2615.1

\item[] Bauer, P. and Thorpe, A. (2015). The Quiet Revolution of Numerical 
Weather Prediction,
\textit{Nature} \textbf{525}: 47--55.
%doi:10.1038/nature14956
